---
title: "Gaussian Elimination in the matrixDemos Package"
author: "John Fox"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r echo=FALSE, message=FALSE}
# include this code chunk as-is to set options
knitr::opts_chunk$set(comment=NA, prompt=TRUE)
library(matrixDemos)
```

Gaussian elimination figures prominently in the **matrixDemos** package, and is implemented directly in the `GaussianElimination` function. Gaussian elimination is also used by `matrixInverse` to compute the inverse of a nonsingular square matrix; by `Ginv` to compute a generalized inverse of a singular square matrix or a rectangular matrix; and by `RREF` to compute the reduced row-echelon form of a matrix.

`GaussianElimination` has one required argument, the matrix `A`, which is transformed to reduced row-echelon form; for example:
```{r}
(A <- matrix(c(1,2,3,4,5,6,7,8,10), 3, 3)) # a nonsingular matrix
GaussianElimination(A)

(B <- matrix(1:9, 3, 3)) # a singular matrix
GaussianElimination(B)
```
In these applications, `GaussianElimination` is equivalent to `RREF`.

A second optional argument to `GaussianElimination` is `B`, which may be a matrix or a vector, and which is transformed in the same manner as `A`. For example,
```{r}
A # nonsingular
GaussianElimination(A, diag(3)) # find inverse of A

b <- 1:3
GaussianElimination(A, b)  # solve the matrix equation Ax = b
```
For the last example, the solution is $x_1 = 1, x_2 = x_3 = 0$.

Other arguments to `GaussianElimination`:

* `tol`, allows you to control the *tolerance* for checking that a number is (effectively) 0 within rounding error

* `verbose`, if `TRUE` (default is `FALSE`) shows the process of reducing `A` to RREF

* `fractions`, if `TRUE` (default is `FALSE`) tries to express the result as rational (rather than decimal) fractions

Some further examples:
```{r}
matrixInverse(A, fractions=TRUE)
RREF(A, verbose=TRUE)
Ginv(B, fractions=TRUE)  # a generalized inverse of singular B
B %*% Ginv(B) %*% B   # check

(C <- matrix(1:12, 3, 4)) # rectangular matrix
Ginv(C, fractions=TRUE)
round(C %*% Ginv(C) %*% C, 5)   # check